---
title: "Assignment04 - Economic Opportunity and Upward Mobility in the United States"
subtitle: "Exploring Economic and Social Conditions Across the U.S. Counties"
author: "Jonas Wenner (jw2558)"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  warning: false
  message: false
---

## Data

**Data Source:**
The data used in this analysis and visualizations come from the 'Urban 
Institute's Upward Mobility Framework'. The data set showcases longitudinal 
county-level measures related to economic and social conditions in relation 
with upward mobility in the US.
(URL:https://datacatalog.urban.org/dataset/mobility-metrics-data-upward-mobility-framework)

```{r}
#| output: false

# Loading the necessary packages
library(tidyverse)
library(ggplot2)

# Loading the data into folder using Assingment02 method
mobility_link <-
  paste0("https://urban-data-catalog.s3.amazonaws.com/drupal-root-live/2026/06/03/11_mobility-metrics_county_longitudinal.csv")

if (!file.exists("11_mobility-metrics_county_longitudinal.csv")) {
  
  download.file(mobility_link,
                 destfile = "11_mobility-metrics_county_longitudinal.csv")
} 

# Loading the data into the global environment
mobility <- read_csv("11_mobility-metrics_county_longitudinal.csv")

# Taking a glimpse into the data
glimpse(mobility)

```

## Visualization 1 - Changes in Employment Across U.S. Counties

```{r}

# Filtering NAs from the variable 'share_employed', calculating the mean by 
# year and visualizing the average employment share over time 
mobility |>
  filter(!is.na(share_employed)) |>
  group_by(year) |>
  summarize(
    mean_share_employed = mean(share_employed)
  ) |>
  ggplot(mapping = aes(x = year, y = mean_share_employed)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
    labels = scales::label_percent()
  ) +
  scale_x_continuous(
    breaks = 2014:2024
  ) +
  labs(
    title = "Average Employment Share Across U.S. Counties",
    subtitle = "Unweighted mean county employment share, 2014-2024",    
    x = "Year", 
    y = "Average employment share",
    caption = "Source: Urban Institute, Upward Mobility Framework. No data available for 2020" 
  ) +
  theme_minimal()

```

**Interpretation:**
The average employment share across U.S. counties increased from ~76% in 2014
to >80% in 2024. The increase remains relatively consistent until 2019, 
followed by a lower share in 2021 and renewed increase until 2024. Since the
measure captures an unweighted county average, the visual highlights changes in the
average rather than the employment share of the total U.S. population. While
it suggests an improving employment siutation, it cannot explain whether these
opportunities are distributed evenly across counties.


## Visualization 2

```{r}

```

**Interpretation:**

## Visualization 3

```{r}

```

**Interpretation:**

## Visualization 4
```{r}

```

**Interpretation:**

## Conclusion 

