---
title: "Assignment04 - Economic Opportunity and Upward Mobility in the United States"
subtitle: "Exploring Economic and Social Conditions Across the U.S. Counties"
author: "Jonas Wenner (jw2558)"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  warning: false
  message: false
---

## Data

**Data Source:**
The data used in this analysis and visualizations come from the 'Urban 
Institute's Upward Mobility Framework'. The data set showcases longitudinal 
county-level measures related to economic and social conditions in relation 
with upward mobility in the US.
(URL:https://datacatalog.urban.org/dataset/mobility-metrics-data-upward-mobility-framework)

```{r}
#| output: false

# Instruction to not use using scientific numbers
options(scipen = 999)

# Loading the necessary packages
library(tidyverse)
library(ggplot2)

# Loading the data into folder using Assingment02 method
mobility_link <-
  paste0("https://urban-data-catalog.s3.amazonaws.com/drupal-root-live/2026/06/03/11_mobility-metrics_county_longitudinal.csv")

if (!file.exists("11_mobility-metrics_county_longitudinal.csv")) {
  
  download.file(mobility_link,
                 destfile = "11_mobility-metrics_county_longitudinal.csv")
} 

# Loading the data into the global environment
mobility <- read_csv("11_mobility-metrics_county_longitudinal.csv")

# Taking a glimpse into the data
glimpse(mobility)

```

## Visualization 1 - Changes in Employment Across U.S. Counties

```{r}

# Filtering NAs from the variable 'share_employed', calculating the mean by 
# year and visualizing the average employment share over time, with a line graph
mobility |>
  filter(!is.na(share_employed)) |>
  group_by(year) |>
  summarize(
    mean_share_employed = mean(share_employed)
  ) |>
  ggplot(mapping = aes(x = year, y = mean_share_employed)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = 2014:2024
  ) +
  labs(
    title = "Average Employment Share Across U.S. Counties",
    subtitle = "Unweighted mean county employment share, 2014-2024",    
    x = "Year", 
    y = "Average employment share (proportion)",
    caption = "Source: Urban Institute, Upward Mobility Framework. No data available for 2020." 
  ) +
  theme_minimal()

```

**Interpretation:**
The average employment share across U.S. counties increased from ~76% in 2014
to >80% in 2024. The increase remains relatively consistent until 2019, 
followed by a lower share in 2021 and renewed increase until 2024. Since the
measure captures an unweighted county average, the visual highlights changes in 
the average rather than the employment share of the total U.S. population. 
While it suggests an improving employment situation, it cannot explain whether 
these opportunities are distributed evenly across counties. Hence, policymakers 
need to consider geographic differences to identify counties that may not have 
equally benefited from the increase, instead of only relying on the overall 
average.


## Visualization 2 - Differences in Household Income Across U.S. Counties

```{r}

# Filtering for 2024 and non-missing pctl_income_50 values and visualizing
# the distribution of household incomes with a histogram
mobility |>
  filter(year == 2024,
         !is.na(pctl_income_50)
         ) |>
  ggplot(mapping = aes(x = pctl_income_50)) +
  geom_histogram(binwidth = 5000, fill = "orange") +
  labs(
    title = "Distribution of Household Income Across U.S. Counties",
    subtitle = "Household income at the 50th percentile, 2024",
    x = "Household income at the 50th percentile (USD)",
    y = "Number of counties",
    caption = "Source: Urban Institute, Upward Mobility Framework." 
  ) +
  theme_minimal()
  
```

**Interpretation:**
In 2024, household income varied considerably across U.S. counties, while the 
histogram suggests most counties being concentrated around the middle of the 
distribution at around 60-70k. Importantly, the distribution is right-skewed, 
with a small number of counties showcasing a substantial higher 50th-percentile 
household income compared to most other counties. Even though having seen the 
first visualization of the overall improvement in employment, this histogram 
suggests that economic conditions differ considerably across counties. These 
geographic differences suggest that general national trends alone may hide 
important differences in local economic opportunities for citizens. For 
policymakers, this showcases the necessity to consider local economic 
conditions whenever they design policies aimed at improving economic 
opportunities.


## Visualization 3 - Employment and Household Income Across U.S. Counties

```{r}

# Filtering for 2024 and non-missing 'shared_employed' and 'pctl_income_50' 
# values, then visualizing their relationship 
mobility |>
  filter(
    year == 2024,
    !is.na(share_employed),
    !is.na(pctl_income_50)
  ) |>
  ggplot(mapping = aes(x = share_employed, y = pctl_income_50)) +
  geom_point(mapping= aes(
    size = population,
    color = share_employed_quality,
    shape = pctl_income_50_quality), 
    alpha = 0.4) +
  geom_smooth(method = "lm") +
  scale_size_continuous(
    breaks = c(2500000, 5000000, 7500000)) +
  labs(
    title = "Employment and Household Income Across U.S. Counties",
    subtitle = "Employment-to-population ratio and 50th-percentile household income, 2024",
    x = "Employment-to-population ratio (ages 25-54, proportion)",
    y = "Household income at the 50th percentile (USD)",
    color = "Employment data quality",
    shape = "Income data quality",
    size = "County population",
    caption = "Source: Urban Institute, Upward Mobility Framework." 
  ) +
  theme_minimal()

```

**Interpretation:**
The scatterplot highlights that in 2024 a positive relationship between 
employment and household income exists across U.S. counties. Counties with 
higher employment-to-population ratios seem to have higher household incomes at 
the 50th percentile. Nevertheless, there is variation in household incomes when 
comparing counties with similar employment levels, thereby indicating that 
solely employment cannot explain varying economic outcomes. Point size 
represents county population, while color and shape indicate the data quality 
of the employment and income measures. While the relationship cannot be 
interpreted as causal, it suggests that policies targeting employment 
opportunities need to consider other local factors, that may shape incomes. 


## Visualization 4 - Housing Affordability Across Income Levels
```{r}

housing_longer <- pivot_longer(
  data = mobility,
  cols = c(
    share_affordable_30_ami,
    share_affordable_50_ami,
    share_affordable_80_ami
  ),
  names_to = "ami_level",
  values_to = "affordability"
)

housing_longer |>
  filter(
    year == 2024,
    !is.na(affordability)
  ) |>
  group_by(ami_level) |>
  summarize(
    mean_affordability = mean(affordability)
  ) |>
  ggplot(mapping = aes(x = ami_level, y = mean_affordability, fill = ami_level)) +
  geom_col() +
  labs(
    title = "Housing Affordability Across Income Levels",
    subtitle = "Average county affordability ratio by AMI level, 2024",
    x = "Area Median Income (AMI) level",
    y = "Average housing affordability ratio",
    fill = "AMI level",
    caption = "Source: Urban Institute, Upward Mobility Framework."
  ) + 
  theme_minimal()
```

**Interpretation:**
The barplot shows the differences in the average housing affordability ratio across varying income levels in U.S. counties in 2024. The shown ratio is lowest for households at the 30% AMI level, increases at 50% AMI level, and slightly decreases at 80% AMI levels. This suggests rather than following the same pattern for all income groups, housing affordability conditions are different depending on household income levels. For instance, the lower ratio at 30% AMI level highlights that the lowest-income households are facing different housing affordability conditions than households at higher levels. For policymakers, these variations are crucial to determine policies aiming at improving access to affordable housing, while considering different income levels. 


## Conclusion 
Overall, the visualizations highlight that economic opportunity varies greatly across U.S. counties and are impossible to be capture by one single indicator. While average employment has increased over the past years, great differences in household income remain across counties, and employment as an indicator cannot explain these differences alone. In addition, housing affordability conditions also highlighted variation across different income levels, hence adding another dimension for measuring and determine varying levels of economic opportunity for researchers and policymakers. Together, the visualizations underline the importance of considering a wide variety of local economic and social conditions when designing and passing policies aimed at improving upward mobility.
