---
title: "Assignment04 - Economic Opportunity and Upward Mobility in the United States"
subtitle: "Exploring Economic and Social Conditions Across the U.S. Counties"
author: "Jonas Wenner (jw2558)"
format: 
  html:
    embed-resources: true
editor: source
execute: 
  warning: false
  message: false
---

## Data

**Data Source:**
The data used in this analysis and visualizations come from the 'Urban 
Institute's Upward Mobility Framework'. The data set showcases longitudinal 
county-level measures related to economic and social conditions in relation 
with upward mobility in the US.
(URL:https://datacatalog.urban.org/dataset/mobility-metrics-data-upward-mobility-framework)

```{r}
#| output: false

options(scipen = 999)

# Loading the necessary packages
library(tidyverse)
library(ggplot2)

# Loading the data into folder using Assingment02 method
mobility_link <-
  paste0("https://urban-data-catalog.s3.amazonaws.com/drupal-root-live/2026/06/03/11_mobility-metrics_county_longitudinal.csv")

if (!file.exists("11_mobility-metrics_county_longitudinal.csv")) {
  
  download.file(mobility_link,
                 destfile = "11_mobility-metrics_county_longitudinal.csv")
} 

# Loading the data into the global environment
mobility <- read_csv("11_mobility-metrics_county_longitudinal.csv")

# Taking a glimpse into the data
glimpse(mobility)

```

## Visualization 1 - Changes in Employment Across U.S. Counties

```{r}

# Filtering NAs from the variable 'share_employed', calculating the mean by 
# year and visualizing the average employment share over time 
mobility |>
  filter(!is.na(share_employed)) |>
  group_by(year) |>
  summarize(
    mean_share_employed = mean(share_employed)
  ) |>
  ggplot(mapping = aes(x = year, y = mean_share_employed)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
    labels = scales::label_percent()
  ) +
  scale_x_continuous(
    breaks = 2014:2024
  ) +
  labs(
    title = "Average Employment Share Across U.S. Counties",
    subtitle = "Unweighted mean county employment share, 2014-2024",    
    x = "Year", 
    y = "Average employment share",
    caption = "Source: Urban Institute, Upward Mobility Framework. No data available for 2020" 
  ) +
  theme_minimal()

```

**Interpretation:**
The average employment share across U.S. counties increased from ~76% in 2014
to >80% in 2024. The increase remains relatively consistent until 2019, 
followed by a lower share in 2021 and renewed increase until 2024. Since the
measure captures an unweighted county average, the visual highlights changes in 
the average rather than the employment share of the total U.S. population. 
While it suggests an improving employment situation, it cannot explain whether 
these opportunities are distributed evenly across counties. Hence, policymakers 
need to consider geographic differences to identify counties that may not have 
equally benefited from the increase, instead of only relying on the overall 
average.


## Visualization 2 - Differences in Household Income Across U.S. Counties

```{r}

# Filtering for 2024 and non-missing pctl_income_50 values and visualizing
# the distribution of household incomes across U.S. counties
mobility |>
  filter(year == 2024,
         !is.na(pctl_income_50)
         ) |>
  ggplot(mapping = aes(x = pctl_income_50)) +
  geom_histogram(binwidth = 5000, fill = "orange") +
  labs(
    title = "Distribution of Household Income Across U.S. Counties",
    subtitle = "Household income at the 50th percentile, 2024",
    x = "Household income at the 50th percentile (USD)",
    y = "Number of counties",
    caption = "Source: Urban Institute, Upward Mobility Framework." 
  ) +
  theme_minimal()
  
```

**Interpretation:**
In 2024, household income varied considerably across U.S. counties, while the 
histogram suggests most counties being concentrated around the middle of the 
distribution at around 60-70k. Importantly, the distribution is right-skewed, 
with a small number of counties showcasing a substantial higher 50th-percentile 
household income compared to most other counties. Even though having seen the 
first visualization of the overall improvement in employment, this histogram 
suggests that economic conditions differ considerably across counties. These 
geographic differences suggest that general national trends alone may hide 
important differences in local economic opportunities for citizens. For 
policymakers, this showcases the necessity to consider local economic 
conditions whenever they design policies aimed at improving economic 
opportunities.


## Visualization 3 - Employment and Household Income Across U.S. Counties

```{r}

#
mobility |>
  filter(
    year == 2024,
    !is.na(share_employed),
    !is.na(pctl_income_50)
  ) |>
  ggplot(mapping = aes(x = share_employed, y = pctl_income_50)) +
  geom_point(mapping= aes(size = population), alpha = 0.4) +
  geom_smooth(method = "lm") +
  scale_x_continuous(
    labels = scales::label_percent()
  ) +
  labs(
    title = "Employment and Household Income Across U.S. Counties",
    subtitle = "Employment-to-population ratio and 50th-percentile household income, 2024",
    x = "Employment-to-population ratio (ages 25-54)",
    y = "Household income at the 50th percentile (USD)",
    caption = "Source: Urban Institute, Upward Mobility Framework" 
  ) +
  theme_minimal()

```

**Interpretation:**
The scatterplot highlights that in 2024 a positive relationship between 
employment and household income exists across U.S. counties. Counties with 
higher employment-to-population seem to have higher household incomes at the 
50th percentile. Nevertheless, there is variation in household incomes when 
comparing counties with similar employment levels, thereby indicating that 
solely employment cannot explain varying economic outcomes. While the 
relationship cannot be interpreted as causal, it suggests that policies 
targeting employment opportunities need to consider other local factors, that 
may shape incomes. 


## Visualization 4
```{r}

```

**Interpretation:**

## Conclusion 

